Windowed least-squares model reduction for dynamical systems

### 2019-10-18 section4/spatial_numerical.tex
update

\subsubsection{Linear multistep schemes}

@@

The practical implementation of the Gauss--Newton algorithm requires an
efficient method for computing the Jacobian of the residual over the time
window
$\frac{\partial \residLMSSlabArg{n}}{\partial \genstatecollocMatySlabArg{}}$.
For this purpose, we can leverage the fact that this Jacobian is a block lower
triangular matrix with the following sparsity pattern (for $\lmsWidthArg{n}{i} = 1$,
$i=1,\ldots,\nstepsArg{n}$): 
\begin{equation*} \begin{bmatrix*}[l]
\matshapea & \\%[-5pt]
 \matshapea & \matshapea & \\%[-5pt]
 & \matshapea  & \matshapea & \\%[-5pt]
&  & \ddots & \\%[-5pt]
 & &  & \matshapea &  \matshapea 
\end{bmatrix*},
% \begin{bmatrix*}[c]
%\matshapeb & \\[-5pt]
% & \hspace{-12pt}\matshapea & \\[-5pt]
% &  & \hspace{-12pt}\matshapea & \\[-5pt]
%&  & \hspace{5pt} \ddots & \\[-5pt]
% & &  & & \hspace{-15 pt} \matshapea 
%\end{bmatrix*}.
\end{equation*}
where each block comprises an $\fomdim \times \romdimArg{n}$ dense matrix.
In particular, solution techniques can leverage this structure, e.g., to
efficiently compute Jacobian--vector products.
Another consequence of this sparsity pattern is that the normal equations
arising at each Gauss--Newton iteration 
comprise a banded block system that can also be exploited.

@@

\begin{algorithm} \caption{\spatialAcronym\ trial subspace: algorithm for the FBSM over the $n$th window.} \label{alg:st_iter} 
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; relaxation factor, $\rho \le 1$; growth factor, $\fbsmGrowth \ge 1$; decay
factor, $\fbsmDecay \ge 1$} 
\Output{Stationary point, $\genstateArgnt{n}$ }
\textbf{Online Steps:}\\ 
$\genstate^n_0 \leftarrow \bz$ \Comment{Set initial guess for state} \\
$\controllerArgnt{n} \leftarrow \boldsymbol 0$ \Comment{Set initial
guess for costate}\\ 
$\text{Compute } \genstateArgnt{n}_1 \text{ satisfying }  \massArg{n} \genstateDotArgnt{n}_1(t)  -  \basisspaceTArg{n} \stweightingMatArg{n}
\velocity(\veloargsromArg{1}) =  \massArg{n} \controllerArg{n}{t}$ 
% Eq.~\ref{eq:lspg_continuous} with $\adjoint^n(t) = \controllerArg{n}{t}$ to
% obtain $\genstate_0^{n}(t)$ 
\Comment{Solve~\eqref{eq:lspg_continuous}}\\ 
%Set: $\genstate^n_1 = \genstate$ \Comment{Set state} \\
%$\objectiveArg{n}({\genstate_1^n(t)})$
%\Comment{Evaluate objective function}\\
$i \leftarrow 1$ \Comment{Set counter}\\
\While{$\epsilon \le \int_{\timeStartArg{n}}^{\timeEndArg{n}} \norm{\genstate^n_{i}(t) - \genstate^n_{i-1}(t) }dt $}{
\small{
\begin{multline*}
\text{ Compute } \adjointArgnt{n} \text{ satisfying }
\massArg{n} \adjointDotArg{n}{t}  + \basisspaceTArg{n} \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\basisspaceArg{n} \genstate^n_i(t) + \stateInterceptArg{n},t) \bigg]^T \stweightingMatArg{n} \basisspaceArg{n} \adjointArg{n}{t}= \\ -\bigg[\basisspaceTArg{n} \bigg[ \frac{\partial \velocity}{\partial \stateyDiscrete} ( \basisspaceArg{n} \genstate^n_i(t) + \stateInterceptArg{n},t) \bigg]^T \stweightingMatArg{n} \bigg( \mathbf{I} -   \basisspaceArg{n} [\massArg{n}]^{-1} \basisspaceTArg{n} \stweightingMatArg{n} \bigg)  \bigg( \basisspaceArg{n} \dot{\genstate}_i^n(t)   -   \velocity( \basisspaceArg{n} \genstate_i^n(t)  +\stateInterceptArg{n},t) \bigg) \bigg] 
\end{multline*} }
\Comment{Solve~\eqref{eq:lspg_adjoint} to obtain guess to costate} \\
%Solve Eq.~\ref{eq:lspg_adjoint} with $\genstate^n(t) = \genstate^n_i(t)$ to obtain $\adjoint^n(t)$ \\
$\controllerArgnt{n}  \leftarrow \rho \controllerArgnt{n} + (1 - \rho) \adjointArgnt{n}$ \Comment{Weighted update to costate}\\
$i \leftarrow i+1$ \Comment{Update counter}\\
$\text{Compute }\genstateArgnt{n}_i \text{ satisfying } \massArg{n} \genstateDotArgnt{n}_i(t)   -  \basisspaceTArg{n} \stweightingMatArg{n} \velocity(\basisspaceArg{n} \genstate^n_i(t) + \stateInterceptArg{n},t) = \massArg{n} \controllerArg{n}{t} $
\Comment{Solve~\eqref{eq:lspg_continuous}}\\
%$\objectiveArg{n}({\genstate_i^n(t)})$ \Comment{Evaluate objective function}\\
\uIf{ $\objectiveArg{n}({\basisspaceArg{n}\genstate_i^n + \stateInterceptArg{n}\otimes \onesFunctionArg{n}}) \le \objectiveArg{n}({\basisspaceArg{n}\genstate_{i-1}^n + \stateInterceptArg{n}\otimes \onesFunctionArg{n}})$}
{
$\rho \leftarrow \text{min}(\rho \fbsmGrowth,1)$ \Comment{Grow the relaxation factor}\\
}
\Else{
$\rho \leftarrow \frac{\rho }{ \fbsmDecay}$ \Comment{Shrink the relaxation factor}\\ 
$\genstate_i^{n} \leftarrow  \genstate_{i-1}^{n}$ \Comment{Reset state to value at previous iteration}
}
}
Return converged solution, $\genstateArgnt{n}= \genstateArgnt{n}_i$
\end{algorithm}
\subsubsection{Numerical considerations for the forward and backward problems}
The FBSM requires solving the systems~\eqref{eq:lspg_continuous} and~\eqref{eq:lspg_adjoint}, both of which are defined at the time-continuous level. 
The numerical implementation of the FBSM requires two main ingredients: (1) temporal discretization schemes for the forward and backward problems and (2) 
an efficient method for computing the action of the Jacobian transpose on a vector.

This work examines the use of linear multistep methods for the purpose of temporal discretization of the forward and 
backwards problems. As described in  
the direct solution approach, temporal discretization is achieved by introducing $\nstepsArg{n} + 1$ time instances over each time window. Linear multistep methods then use this time-grid for temporal discretization of~\eqref{eq:lspg_continuous} and~\eqref{eq:lspg_adjoint}.

The second ingredient, namely an efficient method for computing the action of the Jacobian transpose on a vector, can be challenging. 
In the case one does not have access to the full-order model Jacobian (or it is too costly to compute), the evaluation of this term can be challenging and expensive. Here, two methods are discussed that can be used to evaluate the action of the Jacobian transpose on a vector as seen in~\eqref{eq:lspg_adjoint}:
\begin{enumerate}
\item \textit{Jacobian-free approximation}: A non-intrusive way to evaluate the Jacobian transpose in~\eqref{eq:lspg_adjoint} is to recognize that all terms including the Jacobian transpose are left multiplied by the transpose of the trial basis. One can make the manipulation,
$$\basisspaceTArg{n} \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn)\bigg]^T = \bigg[  \frac{\partial \velocity}{\partial \stateyDiscrete} (\veloargsromn) \basisspaceArg{n} \bigg]^T.$$
This manipulation allows one to compute the action of the Jacobian on each column of $\basisspaceArg{n}$ by, e.g., the finite difference approximation
$$\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn) \basisvec_i^n \approx \frac{1}{\epsilon}\bigg( \velocity(\basisspaceArg{n}\genstateArg{n}{t} + \stateInterceptArg{n} + \epsilon \basisvec_i^n,t) - \velocity(\veloargsromn) \bigg).$$
The Jacobian transpose can be formed in $\romdimArg{n}+1$ evaluations of the velocity. For cases where the reduced-order model is low dimensional, this approach is feasible. For higher dimensional reduced-order models, this approach can be prohibitively expensive.

\item \textit{Automatic differentiation}: A more intrusive, but potentially
	more efficient, method of computing the action of the Jacobian transpose on
		a vector is through automatic differentiation (AD). AD methods comprise a
		class of techniques that can be used to numerically evaluate derivatives
		of functions (e.g., Jacobians, vector-Jacobian products) by recursively
		applying the chain rule. The numerical examples presented later in this
		work leverage AD. The principal drawback of AD is that AD methods are intrusive and may not be suitable for, e.g., legacy codes.  
\end{enumerate}

\begin{remark}\label{remark:fbsm}(Acceleration of Indirect Methods)
The FBSM is a simple iterative method for solving the coupled two-point boundary value problem. For large time windows, however, the FBSM may require many 
forward--backward iterations for convergence. More sophisticated solution techniques, such as a multiple FBSM method or multiple shooting methods, promise 
to reduce this cost. Analyses of additional solution techniques will be a subject of future work.
\end{remark}

>>>>

\subsubsection{Linear multistep schemes}\label{sec:lmm}

@@

The practical implementation of the Gauss--Newton algorithm requires an
efficient method for computing the Jacobian of the residual over the time
window
${\partial \residLMSSlabArg{n}}/{\partial \genstatecollocMatySlabArg{}}$.
For this purpose, we can leverage the fact that this Jacobian is a block lower
triangular matrix with the following sparsity pattern (for $\lmsWidthArg{n}{i} = 1$,
$i=1,\ldots,\nstepsArg{n}$): 
\begin{equation*} \begin{bmatrix*}[l]
\matshapea & \\%[-5pt]
 \matshapea & \matshapea & \\%[-5pt]
 & \matshapea  & \matshapea & \\%[-5pt]
&  & \ddots & \\%[-5pt]
 & &  & \matshapea &  \matshapea 
\end{bmatrix*},
% \begin{bmatrix*}[c]
%\matshapeb & \\[-5pt]
% & \hspace{-12pt}\matshapea & \\[-5pt]
% &  & \hspace{-12pt}\matshapea & \\[-5pt]
%&  & \hspace{5pt} \ddots & \\[-5pt]
% & &  & & \hspace{-15 pt} \matshapea 
%\end{bmatrix*}.
\end{equation*}
where each block comprises an $\fomdim \times \romdimArg{n}$ dense matrix.
In particular, solution techniques can leverage this structure, e.g., to
efficiently compute Jacobian--vector products.
Another consequence of this sparsity pattern is that the normal equations
arising at each Gauss--Newton iteration 
comprise a banded block system that can also be exploited.

@@

\begin{algorithm} \caption{\spatialAcronym\ trial subspace: algorithm for the FBSM over the $n$th window.} \label{alg:st_iter} 
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tolerance, $\epsilon$; relaxation factor, $\rho \le 1$; growth factor, $\fbsmGrowth \ge 1$; decay
factor, $\fbsmDecay \ge 1$} 
\Output{Stationary point, $\genstateArgnt{n}$ }
\textbf{Online Steps:}\\ 
$\genstate^n_0 \leftarrow \bz$ \Comment{Set initial guess for state} \\
$\controllerArgnt{n} \leftarrow \boldsymbol 0$ \Comment{Set initial
guess for costate}\\ 
$\text{Compute } \genstateArgnt{n}_1 \text{ satisfying }  \massArg{n} \genstateDotArgnt{n}_1(t)  -  \basisspaceTArg{n} \stweightingMatArg{n}
\velocity(\veloargsromArg{1}) =  \massArg{n} \controllerArg{n}{t}$ 
% Eq.~\ref{eq:lspg_continuous} with $\adjoint^n(t) = \controllerArg{n}{t}$ to
% obtain $\genstate_0^{n}(t)$ 
\Comment{Solve~\eqref{eq:lspg_continuous}}\\ 
%Set: $\genstate^n_1 = \genstate$ \Comment{Set state} \\
%$\objectiveArg{n}({\genstate_1^n(t)})$
%\Comment{Evaluate objective function}\\
$i \leftarrow 1$ \Comment{Set counter}\\
\While{$\epsilon \le \int_{\timeStartArg{n}}^{\timeEndArg{n}} \norm{\genstate^n_{i}(t) - \genstate^n_{i-1}(t) }dt $}{
\small{
\begin{multline*}
\text{ Compute } \adjointArgnt{n} \text{ satisfying }
\massArg{n} \adjointDotArg{n}{t}  + \basisspaceTArg{n} \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}(\basisspaceArg{n} \genstate^n_i(t) + \stateInterceptArg{n},t) \bigg]^T \stweightingMatArg{n} \basisspaceArg{n} \adjointArg{n}{t}= \\ -\bigg[\basisspaceTArg{n} \bigg[ \frac{\partial \velocity}{\partial \stateyDiscrete} ( \basisspaceArg{n} \genstate^n_i(t) + \stateInterceptArg{n},t) \bigg]^T \stweightingMatArg{n} \bigg( \mathbf{I} -   \basisspaceArg{n} [\massArg{n}]^{-1} \basisspaceTArg{n} \stweightingMatArg{n} \bigg)  \bigg( \basisspaceArg{n} \dot{\genstate}_i^n(t)   -   \velocity( \basisspaceArg{n} \genstate_i^n(t)  +\stateInterceptArg{n},t) \bigg) \bigg] 
\end{multline*} }
\Comment{Solve~\eqref{eq:lspg_adjoint} to obtain guess to costate} \\
%Solve Eq.~\ref{eq:lspg_adjoint} with $\genstate^n(t) = \genstate^n_i(t)$ to obtain $\adjoint^n(t)$ \\
$\controllerArgnt{n}  \leftarrow \rho \controllerArgnt{n} + (1 - \rho) \adjointArgnt{n}$ \Comment{Weighted update to costate}\\
$i \leftarrow i+1$ \Comment{Update counter}\\
$\text{Compute }\genstateArgnt{n}_i \text{ satisfying } \massArg{n} \genstateDotArgnt{n}_i(t)   -  \basisspaceTArg{n} \stweightingMatArg{n} \velocity(\basisspaceArg{n} \genstate^n_i(t) + \stateInterceptArg{n},t) = \massArg{n} \controllerArg{n}{t} $
\Comment{Solve~\eqref{eq:lspg_continuous}}\\
%$\objectiveArg{n}({\genstate_i^n(t)})$ \Comment{Evaluate objective function}\\
\uIf{ $\objectiveArg{n}({\basisspaceArg{n}\genstate_i^n + \stateInterceptArg{n}\otimes \onesFunctionArg{n}}) \le \objectiveArg{n}({\basisspaceArg{n}\genstate_{i-1}^n + \stateInterceptArg{n}\otimes \onesFunctionArg{n}})$}
{
$\rho \leftarrow \text{min}(\rho \fbsmGrowth,1)$ \Comment{Grow the relaxation factor}\\
}
\Else{
$\rho \leftarrow \frac{\rho }{ \fbsmDecay}$ \Comment{Shrink the relaxation factor}\\ 
$\genstate_i^{n} \leftarrow  \genstate_{i-1}^{n}$ \Comment{Reset state to value at previous iteration}
}
}
Return converged solution, $\genstateArgnt{n}= \genstateArgnt{n}_i$
\end{algorithm}
\subsubsection{Considerations for the numerically solving the forward and
backward systems}
The FBSM requires solving the forward~\eqref{eq:lspg_continuous} and
backward~\eqref{eq:lspg_adjoint} systems, both of which are defined at the time-continuous level. 
The numerical implementation of the FBSM requires two main ingredients: (1) temporal discretization schemes for the forward and backward problems and (2) 
an efficient method for computing terms involving the transpose of the
Jacobian of the velocity.

This work employs linear multistep schemes for time discretization of the
forward and backward problems. As described in Section \ref{sec:lmm},
temporal discretization is achieved by introducing $\nstepsArg{n} + 1$ time
instances over each time window.

The second ingredient, namely devising an efficient method for computing terms
involving the transpose of the Jacobain of the velocity,
can be challenging if one
does not have explicit access to this Jacobian or it is too costly to
compute. We discuss two methods that can be used to evaluate such terms
that appear in the forward system~\eqref{eq:lspg_adjoint}:
\begin{enumerate}
\item \textit{Jacobian-free approximation}: A non-intrusive way to evaluate
	these terms is to recognize that all terms including the transpose of the
		Jacobian of the velocity are left multiplied by the transpose of the
		spatial trial basis; this can be manipulated as
$$\basisspaceTArg{n} \bigg[\frac{\partial \velocity}{\partial \stateyDiscrete}
		(\veloargsromn)\bigg]^T = \bigg[  \frac{\partial \velocity}{\partial
		\stateyDiscrete} (\veloargsromn) \basisspaceArg{n} \bigg]^T,$$
		which exposes the ability to approximate rows of this matrix via
		finite differences, e.g., via forward differences as
$$\frac{\partial \velocity}{\partial \stateyDiscrete}(\veloargsromn)
		\basisvec_i^n \approx \frac{1}{\epsilon}\bigg(
		\velocity(\basisspaceArg{n}\genstateArg{n}{t} + \stateInterceptArg{n} +
		\epsilon \basisvec_i^n,t) - \velocity(\veloargsromn) \bigg),\quad
		i=1,\ldots,\romdimArg{n},$$
which requires $\romdimArg{n}+1$ evaluations of the velocity.

\item \textit{Automatic differentiation}: A more intrusive, but exact
	method for computing these terms is
		through automatic differentiation (AD), which evaluate derivatives of
		functions (e.g., Jacobians, vector-Jacobian products) in a numerically
		exact manner by recursively applying the chain rule. The numerical
		examples presented later in this work leverage AD. The principal drawback
		of this approach is their intrusiveness, which may prevent them from
		practical application, e.g., in legacy codes.
		\KTC{EP: verify this paragraph.}
\end{enumerate}

\KTC{HERE}
\begin{remark}\label{remark:fbsm}(Acceleration of Indirect Methods)
The FBSM is a simple iterative method for solving the coupled two-point boundary value problem. For large time windows, however, the FBSM may require many 
forward--backward iterations for convergence. More sophisticated solution techniques, such as a multiple FBSM method or multiple shooting methods, promise 
to reduce this cost. Analyses of additional solution techniques will be a subject of future work.
\end{remark}
